---
title: 'Most Damaging Weather Events in US (1996-2011)'
output: 
        html_document: 
                keep_md: true
---
## Synopsis
In this report we aim to determine which weather events in the US have had the
greatest impact on both life and limb, as well as economically. To investigate this,
we obtained data from the US National Weather Service for the years 1996 to 
2011. We found that tornadoes are the most leathal and injurious weather event, 
and the third most damaging. Flood is the most economically damaging and
significant in terms of health impact.

## Data Processing

### Requirements
This analysis requires the following packages.
```{r requirements}
require(dplyr)
require(tidyr)
require(ggplot2)
```

### Obtaining Data
We obtained the data from the [National Weather Service](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
on 19 August 2015 as follows. Further details on the data set can be found
by referring to the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).
```{r fetching data}
BZ_FILE_NAME <- 'StormData.csv.bz2'

if (!file.exists(BZ_FILE_NAME)) {
        download.file(
                url = 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
                destfile = BZ_FILE_NAME,
                method = 'curl'
        )
}
```

### Load the data
The raw data is loaded into a data frame table.
```{r loading data, cache=TRUE}
bz_file <- bzfile(BZ_FILE_NAME)
storm_data <- tbl_df(read.csv(bz_file))
```

Initially, we clean the data by removing all weather events that had no
fatalities, injuries, crop damage and property damage. This greatly reduces the
number of unique event types and has no impact on the analysis.

We are only interested in injuries, fatalities, and economic damage, so we
select these columns.
```{r filterting all-zero data}
intermediate_data <- storm_data %>% 
        filter(FATALITIES > 0 | INJURIES > 0 | CROPDMG > 0 | PROPDMG > 0) %>%
        select(event_type = EVTYPE, fatalities = FATALITIES, injuries = INJURIES,
               crop_damage = CROPDMG, crop_damage_exponent = CROPDMGEXP,
               property_damage = PROPDMG, property_damage_exponent = PROPDMGEXP,
               begin_date = BGN_DATE)
```

[According to NOAA](https://www.ncdc.noaa.gov/stormevents/details.jsp), only the
dates from Jan 1996 contain the full 48 weather types. Therefore, we have opted
to remove dates earlier than this.
```{r filtering dates}
intermediate_data <- intermediate_data[
        strptime(intermediate_data$begin_date, "%m/%d/%Y") >= as.POSIXlt("1996-01-01"),
        ]
```

The max date is:
```{r max date}
max(strptime(intermediate_data$begin_date, "%m/%d/%Y"))
```

The min date is:
```{r min date}
min(strptime(intermediate_data$begin_date, "%m/%d/%Y"))
```

Next we look at property and crop damage. We first looked at the existing 
"exponents," which are codes for multiples to apply to the data. According to 
the documentation (in section 2.7), the only defined codes are K, M and B. We 
assume that where they appear in lowercase, they should be uppercase. Apart 
from this, all other codes are undefined, so we remove records with them.
```{r}
intermediate_data$crop_damage_exponent <-
        toupper(intermediate_data$crop_damage_exponent)
intermediate_data$property_damage_exponent <-
        toupper(intermediate_data$property_damage_exponent)

intermediate_data <- intermediate_data %>%
        filter(crop_damage_exponent %in% c("K", "M", "B")) %>%
        filter(property_damage_exponent %in% c("K", "M", "B"))
```

Further cleaning the data involves multiplying the damage by the values for the
codes. The values are "K" -> 1,000, "M" -> 1,000,000 and "B" -> 1,000,000,000. 
The documentation is ambiguous as to what no code means, therefore, we assume it
is in dollars and use a multiple of 1.

```{r}
intermediate_data$crop_damage <- intermediate_data$crop_damage * 
        ifelse(
                intermediate_data$crop_damage_exponent == "B", 10^9, ifelse(
                        intermediate_data$crop_damage_exponent == "M", 10^6, ifelse(
                                intermediate_data$crop_damage_exponent == "K", 10^3, 1
        )))

intermediate_data$property_damage <- intermediate_data$property_damage * 
        ifelse(
                intermediate_data$property_damage_exponent == "B", 10^9, ifelse(
                        intermediate_data$property_damage_exponent == "M", 10^6, ifelse(
                                intermediate_data$property_damage_exponent == "K", 10^3, 1
        )))
```

We now combine the damage values into one and remove unneeded columns.
```{r}
intermediate_data <- intermediate_data %>% 
        mutate(total_damage = crop_damage + property_damage) %>%
        select(-property_damage, 
               -crop_damage, 
               -property_damage_exponent, 
               -crop_damage_exponent,
               -begin_date)
```

We combine injuries and fatalities into one column for the purposes
of processing, and identify them with a key called fatality or injury.
``` {r}
intermediate_data <- intermediate_data %>% 
        gather(health_effect, health_effect_count, fatalities, injuries)
```

Grouping by event type helps reduce the sheer number of observations and gives
us an insight into the events with the greatest impact over the entire period.
```{r}
intermediate_data <- group_by(intermediate_data, event_type, health_effect) %>%
        summarise(health_effect_count = sum(health_effect_count), 
                  total_damage = sum(total_damage))
```

From the unique event types, we see that there are many potential duplicates:
```{r unique events}
unique(intermediate_data$event_type)
```

To clean this, we first standardize the case of the text:
```{r standardize case}
intermediate_data$event_type <- toupper(intermediate_data$event_type)
```

Next, we apply a mapping to reduce the duplicated names. Note that it is quite
arguable which events should be combined. Our approach was to match the list 
given in section 2.1.1 of the Storm Data documentation.

```{r mapping data}
mapping <- c("(?<=\\w)ING(?=\\s|$|\\W)", "",
             "(?<=\\w)S(?=\\s|$|\\W)", "",
             "^ASTRONOMICAL HIGH TIDE", "STORM SURGE/TIDE",
             "DEBRI FLOW", "DEBRIS FLOW",
             "DRY MICROBURST", "THUNDERSTORM WIND",
             "DUST STORM/HIGH WIND", "DUST STORM",
             "^EXTREME COLD$", "EXTREME COLD/WIND CHILL",
             ".*FLASH FLOOD.*", "FLASH FLOOD",
             "^FOG$", "DENSE FOG",
             "^FOREST FIRE$", "WILDFIRE",
             "^FREEZ FOG$", "FROST/FREEZE",
             "^FREEZE$", "FROST/FREEZE",
             "GLAZE ICE", "FROST/FREEZE",
             "GUSTNADO", "THUNDERSTORM WIND",
             "GUSTY WIND", "HIGH WIND",
             "HAIL 100", "HAIL",
             "HAIL/WIND", "HIGH WIND",
             "^HEAT WAVE$", "HEAT",
             "^HEAT WAVE DROUGHT$", "DROUGHT",
             "HEAVY RAIN/FLOOD", "FLOOD",
             "HEAVY SNOW/HIGH WIND & FLOOD", "FLOOD",
             "HEAVY RAIN/HIGH SURF", "HEAVY RAIN",
             "HEAVY SURF/HIGH SURF", "HIGH SURF",
             "HIGH WIND HEAVY RAIN", "HIGH WIND",
             "HIGH WIND/COLD", "HIGH WIND",
             "^HURRICANE$", "HURRICANE (TYPHOON)",
             "HURRICANE.+", "HURRICANE (TYPHOON)",
             "ICE JAM FLOOD", "FLOOD",
             "ICY ROAD", "FROST/FREEZE",
             "LANDSLIDE", "DEBRIS FLOW",
             "LIGHTN", "LIGHTNING",
             "RIVER FLOOD", "FLOOD",
             "SMALL HAIL", "HAIL",
             "SEVERE THUNDERSTORM.*", "THUNDERSTORM WIND",
             "^STORM SURGE$", "STORM SURGE/TIDE",
             "THUNDERSTORM HAIL", "THUNDERSTORM WIND",
             "THUNDERSTORM WIND.*", "THUNDERSTORM WIND",
             "^THUNDERSTORM$", "THUNDERSTORM WIND",
             "^TORNADO.*", "TORNADO",
             "TROPICAL STORM.+", "TROPICAL STORM",
             "TSTM WIND.*", "THUNDERSTORM WIND",
             "^TYPHOON$", "HURRICANE (TYPHOON)",
             "URBAN FLOOD", "FLOOD",
             "URBAN/SML STREAM FLD", "FLOOD",
             "WIND DAMAGE", "HIGH WIND",
             "^WIND$", "HIGH WIND",
             "WILD/FOREST FIRE", "WILDFIRE",
             "WINTER STORM HIGH WIND", "WINTER STORM"
             )

dim(mapping) <- c(2, length(mapping)/2)

for(i in 1:dim(mapping)[2]) {
        intermediate_data$event_type <- gsub(mapping[1,i], 
                                            mapping[2,i], 
                                            intermediate_data$event_type,
                                            perl = TRUE)
}


sort(unique(intermediate_data$event_type))
```

## Results
We look at the top ten injury causing events and the top ten fatality causing
events. This is because our visualizations would be too dense, and not
particularly useful if we listed all 48 event types.
```{r top ten health effect}
health_effect_type_count <- count(intermediate_data, health_effect)
by_event_type_and_health_effect <- intermediate_data %>%
        group_by(health_effect) %>%
        filter(rank(health_effect_count, ties.method = "first") >
                       health_effect_type_count$n[1] - 10)
```

Plotting these provide insight into which events have the greatest impact in 
terms of health.
```{r, fig.height=7}
               
ggplot(by_event_type_and_health_effect) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        aes(x = event_type, y = health_effect_count, fill = health_effect) +
        geom_bar(stat = "identity", position = "dodge") +
        geom_text(aes(label = health_effect_count), size = 4, vjust = 0) +
        facet_grid(health_effect ~ ., scale = "free_y") +
        xlab("Event Type") +
        ylab("Number of Deaths/Injuries") +
        ggtitle("Top 13 Health Impacting Weather Events (1996-2011)")
```  
    
We can see that tornados are the most fatal and most injurious by far. Flash
floods are second most fatal, but regular floods are second most injurious.

Some interesting observations are that the third most injurious, ice storms, 
are actually not lethal, and hurricanes are quite injurious, but not
terribly lethal. Rip currents, on the otherhand, are fourth most fatal, but are
not at all injurious.

### Top Economic Damaging Events
Next we look at the economic impact of the events. Again, we select only the
top ten most damaging events to reduce the density of the plot, as we are only
interested in the most impactful events, anyway.

```{r}
by_event_type <- group_by(intermediate_data, event_type) %>%
        summarise(total_damage = sum(total_damage)) 

event_type_count <- tally(by_event_type)

top_ten_total_damage_by_event_type <- by_event_type %>%
        filter(rank(total_damage, ties.method = "first") >
                       event_type_count$n[1] - 10)
        
```

```{r, fig.height=6, }
ggplot(top_ten_total_damage_by_event_type) + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        aes(x = event_type, y = total_damage/10^6) +
        geom_bar(stat = "identity") +
        geom_text(aes(label = format(round(total_damage/10^6), big.mark = ",")),
                  size = 4, vjust = -0.5) +
        xlab("Event Type") +
        ylab("Total Damage (millions)") +
        ggtitle("Top 10 Economically Impacting Weather Events (1996-2011)")
```

By far, floods are the most economically damaging events, by an order of 
magnitude. Second are hurricanes, and third are tornados.

It is interesting to compare how much impact an event has in terms of economic 
damage vs health effects. For example, tornados have most impact on human
health, but a distant third in terms of economic damage. This might be due to
the localized but sudden nature of a tornado (localized might mean it has 
trouble doing significant damage, but its suddeness might meant that people
are less able to get out of the way).
